Model Finding for (Co)recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-05 tex/main.tex
more

\newcommand{\pnone}{\con{none}}
\newcommand{\ptrue}{\con{true}}
\newcommand{\pfalse}{\con{false}}
\newcommand{\pol}{\con{pol}}

@@

\newcommand{\boolop}{\oplus}
\newcommand{\forallf}[1]{\forall_{\!#1\:}}
\newcommand{\fnull}{\emptyset}
\newcommand{\farg}[1]{a_{#1}}
\newcommand{\vecfarg}[1]{\vec{a}_{#1}}
\newcommand{\fargsort}[1]{\upsilon_{#1}}

@@

Let us fix a signature $\Sigma$, a $\Sigma$-formula
$\varphi$ in definitional form with respect to functions $\sfundefs{\Sigma}$,
whose definitions are admissible.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature
$\extendsig{ \Sigma }$ and a $\extendsig{ \Sigma }$-formula $\psi$ such that
$\varphi$ and $\psi$ are equisatisfiable.
\jb{Equisatisfiability as defined in Sect.\ 2 requires the same $\Sigma$ and
the same $T$ for both formulas, which isn't the case here.}
The intuition behind this translation
is to use an uninterpreted type \ct{to be defined} to abstract the set of
\emph{relevant} tuples for each recursive function $\con{f}$ and restrict the
quantification of the function definition for $\con{f}$ to a variable of this
type. Informally, the relevant tuples $\vec t$ of a function $\con{f}$ are the
ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the
satisfiability of $\varphi$.

@@

In more detail, 
say our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
We extend the signature of $\Sigma$ to a signature $\extendsig{\Sigma}$ containing the following additional definitions.
For each $\con{f} \in \sfundefs{\Sigma}$ of type $\tau_1 \times \ldots \times \tau_n \rightarrow \tau$, 
the signature $\extendsig{\Sigma}$ includes:
\begin{enumerate}
\item[-] an uninterpreted type $\fargsort{\con{f}}$, and
\item[-] a vector of $n$ uninterpreted functions $\farg{\con{f}}^1$, $\ldots$, $\farg{\con{f}}^n$ of type $(\fargsort{\con{f}} \rightarrow \tau_1)$, $\ldots$, $(\fargsort{\con{f}} \rightarrow \tau_n)$.
\end{enumerate}
%The interpretation of uninterpreted sort $\fargsort{\con{f}}$ will denote the elements (tuples) on which the function $\con{f}$ is applied.
%The role of the uninterpreted functions $\farg{\con{f}}^1$, $\ldots$, $\farg{\con{f}}^n$ will be discussed more in the following.

Given the extended signature, we run the procedure $\conv$ in Figure~\ref{fig:encoding} on $\varphi$.
This calls the subprocedure $\conv_0$, which takes two arguments: the term $t$ to translate, and a \emph{polarity} $p$ (either $\ptrue$, $\pfalse$, or $\pnone$).
It returns a pair of the form $( r, D )$, where $r$ is a term of type $\tau$, and $D$ is a set of formulas.
%The role of $D$ is to ensure that the (restricted) function definition for $\con{f}$ includes certain tuples in its domain, which we explain more in the following.
At a high level, the translation modifies $\varphi$ in two ways.
First, it restricts the quantification on function definitions for $\con{f}$ to the corresponding uninterpreted type $\fargsort{\con{f}}$.
Second, it augments $\varphi$ with additional constraints of the form $\exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( z ) \teq \vec t )$
which ensure that this restriction for the defintion $\con{f}$ covers all relevant tuples of terms,
namely the ones for which an $\con{f}$-application exists and is relevant to the satisfiability of the current branch of $\varphi$.

@@

\begin{theorem}
If $\varphi$ be a $\Sigma$-formula that is in definitional form with respect to functions $\sfundefs{\Sigma}$,
and the set of function definitions corresponding to functions in $\sfundefs{\Sigma}$ is admissible,
then $\varphi$ and $\conv( \varphi )$ are equisatisfiable.
\end{theorem}
\begin{proof}
(Sketch)  First, we show that if $\varphi$ has a model $\I$, then $\conv( \varphi )$ has a model $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \ssorts{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$, 
interprets $\fargsort{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \ldots \times \tau_n^\I$,
and each $\farg{\con{f}}^i$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Note that $\J$ satisfies \emph{every} constraint of the form $\exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( z ) \teq \vec t )$,
since by our construction of $\J$ there exists a $z$ such that $z^\J = \vec t^\J$.
By case analysis it follows that $\conv_0$ maintains the invariant that
if $\conv_0( t\typ{\Bool}, \ptrue )$ returns $( r\typ{\Bool}, D )$,
then $\I$ satisfies $t$ if and only if $\J$ satisfies $r$.
Thus, we conclude $\J$ is a model of $\conv( \varphi )$.

>>>>

\newcommand{\pnone}{\con{none}}
\newcommand{\ptrue}{\con{pos}}
\newcommand{\pfalse}{\con{neg}}
\newcommand{\pol}{\con{pol}}

@@

%\newcommand\concret{\con{a}}
%\newcommand\concret{\con{c}}
\newcommand\concret{\gamma} %%% from the abstract interpretation literature

\newcommand{\boolop}{\oplus}
\newcommand{\forallf}[1]{\forall_{\!#1\:}}
\newcommand{\fnull}{\emptyset}
\newcommand{\vecfarg}[1]{\vec{\concret}_{\vthinspace#1}}
\newcommand{\farg}[1]{\concret_{\vthinspace#1}}
\newcommand{\fargx}[2]{\concret^{\vthinspace#2}_{\vthinspace#1}}
\newcommand{\fargsort}[1]{\upsilon_{\vthinspace#1}}

@@

Let us fix a signature $\Sigma$, a $\Sigma$-formula
$\varphi$ in definitional form with respect to functions $\sfundefs{\Sigma}$,
whose definitions are admissible.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature
$\extendsig{ \Sigma }$ and a $\extendsig{ \Sigma }$-formula $\psi$ such that
$\varphi$ and $\psi$ are equisatisfiable.
\jb{Equisatisfiability as defined in Sect.\ 2 requires the same $\Sigma$ and
the same $T$ for both formulas, which isn't the case here.}
The intuition behind this translation
is to use an uninterpreted type \ct{to be defined} to abstract the set of
\relax{relevant} tuples for each recursive function $\con{f}$ and restrict the
quantification of the function definition for $\con{f}$ to a variable of this
type. Informally, the relevant tuples $\vec t$ of a function $\con{f}$ are the
ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the
satisfiability of $\varphi$.

@@

In more detail, 
suppose our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
We extend the signature of $\Sigma$ to a signature $\extendsig{\Sigma}$ containing the following additional definitions.
For each $\con{f} : \tau_1 \times \ldots \times \tau_n \rightarrow \tau \in \sfundefs{\Sigma}$,
the signature $\extendsig{\Sigma}$ includes
\begin{itemize}
\item an uninterpreted \emph{abstract type} $\fargsort{\con{f}}$;
\item $n$ uninterpreted \emph{concretization functions} $\fargx{\con{f}}{1} : \fargsort{\con{f}} \rightarrow \tau_1$, \ldots, $\fargx{\con{f}}{n} : \fargsort{\con{f}} \rightarrow \tau_n$.
\end{itemize}
%The interpretation of uninterpreted sort $\fargsort{\con{f}}$ will denote the elements (tuples) on which the function $\con{f}$ is applied.
%The role of the uninterpreted functions $\farg{\con{f}}^1$, $\ldots$, $\farg{\con{f}}^n$ will be discussed more in the following.

The procedure $\conv$ in Figure~\ref{fig:encoding} translates the $\Sigma$-formula
$\varphi$ into an $\extendsig{\Sigma}$-formula.
This calls the subprocedure $\conv_0$, which takes two arguments:\ the term $t$
to translate and a \emph{polarity} $p$ (either $\ptrue$, $\pfalse$, or
$\pnone$). It returns a pair $( r,\, D )$, where $r$ is a term of
type $\tau$ and $D$ is a set of formulas.
\jb{Explain r and D here briefly.}
%The role of $D$ is to ensure that
%the (restricted) function definition for $\con{f}$ includes certain tuples in
%its domain, which we explain more in the following.

The translation modifies $\varphi$ in two main ways. First, it restricts the
quantification on function definitions for $\con{f}$ to the corresponding
uninterpreted type $\fargsort{\con{f}}$ and inserts appropriate . Second, it augments $\varphi$ with
additional constraints of the form $\exists z\typ{\fargsort{\con{f}}}. (
\vecfarg{\con{f}}( z ) \teq \vec t )$ which ensure that this restriction for
the defintion $\con{f}$ covers all relevant tuples of terms, namely the ones
for which an $\con{f}$-application exists and is relevant to the satisfiability
of the current branch of $\varphi$.

@@

\begin{theorem}
If $\varphi$ be a $\Sigma$-formula that is in definitional form with respect to functions $\sfundefs{\Sigma}$,
and the set of function definitions corresponding to functions in $\sfundefs{\Sigma}$ is admissible,
then $\varphi$ and $\conv( \varphi )$ are equisatisfiable.
\end{theorem}
\begin{proof}
(Sketch)  First, we show that if $\varphi$ has a model $\I$, then $\conv( \varphi )$ has a model $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \ssorts{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \ldots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$, 
interprets $\fargsort{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \ldots \times \tau_n^\I$,
and each $\fargx{\con{f}}{i}$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Note that $\J$ satisfies \emph{every} constraint of the form $\exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( z ) \teq \vec t )$,
since by our construction of $\J$ there exists a $z$ such that $z^\J = \vec t^\J$.
By case analysis it follows that $\conv_0$ maintains the invariant that
if $\conv_0( t\typ{\Bool}, \ptrue )$ returns $( r\typ{\Bool}, D )$,
then $\I$ satisfies $t$ if and only if $\J$ satisfies $r$.
Thus, we conclude $\J$ is a model of $\conv( \varphi )$.
